What is happening on Twitter? A framework for student research projects with tweets

### 2020-11-02 Rmd/jse-special-issue-2020-november.Rmd
added fixes for November 2020 typos

Some social media data, including tweets from Twitter, is
available through website application product interfaces (APIs). 
By way of a streaming API, Twitter shares a sample of approximately 
one percent of all 
tweets during an API query time period [@tweet_stream]. Any Twitter 
user can freely 
access this one percent sample, whereas access to a larger selection is available to 
researchers for a fee.

@@

```{r, circle-table, echo = FALSE, message=FALSE}
library(magrittr)
readr::read_csv("../data/concepts-backward-design.csv") %>%
  knitr::kable(caption = "Classifying project skills") %>%
  kableExtra::column_spec(column = 1, border_left = TRUE) %>%
  kableExtra::column_spec(column = 2, border_right = TRUE)
```

@@

Our first student examined relationships over time between 
stock market index prices and tweet sentiment. For each day in her 12-month study period, 
she identified stock market-related tweets with a key word search. With the complete
texts of stock market-related tweets for each day, she calculated a daily sentiment score and plotted it over time. 
Her sentiment score reflected presence of emotion-associated terms 
(*eg*., "happy", "sad", "mad", "scared") 
in tweet texts. Days with more net positive emotion words in the collected tweets received a higher 
(positive) daily sentiment score, while days with more net negative words received a negative daily 
sentiment score. 
For her final project, she presented plots over time of her daily sentiment scores and daily closing 
prices of the Standard and Poor's 500 index. She also explored time series analysis methods to quantify 
relationships between index prices and sentiment scores.

Our second student developed social media event detection methods with topic 
models. He hypothesized that tweet content changes over time, and that we 
might detect these changes by comparing inferred tweet topics from distinct 
time periods. To validate his hypothesis, he examined tweet content before, during, and after 
the National Football League's Super Bowl game in 2015. He reasoned that 
because the Super Bowl is widely discussed on Twitter, we might detect 
Super Bowl-related topics from tweets sent during the game, but that the football-related
topics would be short-lived in the continuous Twitter stream. We discovered
evidence to support his strategy, and we ultimately presented our findings 
at international and local research meetings. We share below a case study based
on this approach.

@@

Below is an example of JSON for one tweet.

```{bash, eval = FALSE, fig.cap = "JSON example for one tweet", echo = TRUE}
{
  "created_at": "Thu Apr 06 15:24:15 +0000 2017",
  "id_str": "850006245121695744",
  "text": "1\/ Today we\u2019re sharing our vision for the future of the Twitter API platform!",
  "user": {
    "id": 2244994945,
    "name": "Twitter Dev",
    "screen_name": "TwitterDev",
    "location": "Internet",
    "url": "https:\/\/dev.twitter.com\/",
    "description": "Your official source for Twitter Platform news, updates & events. 
    Need technical help? Visit https:\/\/twittercommunity.com\/ \u2328\ufe0f 
    #TapIntoTwitter"
  },
  "place": {   
  },
  "entities": {
    "hashtags": [      
    ],
    "urls": [
      {
        "url": "https:\/\/t.co\/XweGngmxlP",
        "unwound": {
          "url": "https:\/\/cards.twitter.com\/cards\/18ce53wgo4h\/3xo1c",
          "title": "Building the Future of the Twitter API Platform"
        }
      }
    ],
    "user_mentions": [     
    ]
  }
}
```

@@

LDA model fitting requires that the corpus be organized as a 
document by term matrix. In a document by term matrix, each row corresponds to a single document (a single 
tweet), and each column is a single term (or word). Each cell contains a 
count (the number of 
occurrences of a term in the specified document). We created a document by 
term matrix with the R function `cast_dtm` from the `tidytext` package.

@@

We identified the top ten most probable terms for each of ten topics in our models 
(Figures \@ref(fig:may24), \@ref(fig:may25), \@ref(fig:may26)). 
We plotted the within-topic word probabilities as bar graphs. We see that topic-specific 
word probabilities seldom exceed 0.05. We also note that some words are heavily weighted in 
multiple topics. This observation complicates semantic topic interpretation. We 
also caution that the results display expletives (that appeared on Twitter) and may 
not be "safe for work".

@@

```{r, may25, fig.cap = "Top terms for LDA model from May 25, 2020 (Memorial Day)"}
knitr::include_graphics("../results/beta-2020-05-25.png")
```

>>>>

Some social media data, including tweets from Twitter, is
available through website application programming interfaces (APIs). 
By way of a streaming API, Twitter shares a sample of approximately 
one percent of all 
tweets during an API query time period [@tweet_stream]. Any Twitter 
user can freely 
access this one percent sample, whereas access to a larger selection is available to 
researchers for a fee.

@@

```{r, circle-table, echo = FALSE, message=FALSE}
library(magrittr)
readr::read_csv("../data/concepts-backward-design.csv") %>%
  dplyr::arrange(Category) %>%
  knitr::kable(caption = "Classifying project skills") %>%
  kableExtra::column_spec(column = 1, border_left = TRUE) %>%
  kableExtra::column_spec(column = 2, border_right = TRUE)
```

@@

Our first student examined relationships over time between 
stock market index prices and tweet sentiment. For each day in her 12-month study period, 
she identified stock market-related tweets with a key word search. With the complete
texts of stock market-related tweets for each day, she calculated a daily sentiment score and plotted it over time. 
Her sentiment score reflected presence of emotion-associated terms 
(*e.g*., "happy", "sad", "mad", "scared") 
in tweet texts. Days with more net positive emotion words in the collected tweets received a higher 
(positive) daily sentiment score, while days with more net negative words received a negative daily 
sentiment score. 
For her final project, she presented plots over time of her daily sentiment scores and daily closing 
prices of the Standard and Poor's 500 index. She also explored time series analysis methods to quantify 
relationships between index prices and sentiment scores.

Our second student developed social media event detection methods with topic 
models. He hypothesized that tweet content changes over time, and that we 
might detect these changes by comparing inferred tweet topics from distinct 
time periods. To validate his hypothesis, he examined tweet content before, during, and after 
the National Football League's Super Bowl game in 2015. He reasoned that 
because the Super Bowl is widely discussed on Twitter, we might detect 
Super Bowl-related topics from tweets sent during the game, but that the football-related
topics would be short-lived in the continuous Twitter stream. We discovered
evidence to support his ideas, and we ultimately presented our findings 
at international and local research meetings. Below, we share a case study on a different, widely discussed topic which is analyzed using an approach similar to that from the Super Bowl tweets.

@@

Below is an example of a tweet in JSON format.

```{bash, eval = FALSE, fig.cap = "JSON format for one tweet", echo = TRUE}
{
  "created_at": "Thu Apr 06 15:24:15 +0000 2017",
  "id_str": "850006245121695744",
  "text": "1\/ Today we\u2019re sharing our vision for the future of the Twitter API platform!",
  "user": {
    "id": 2244994945,
    "name": "Twitter Dev",
    "screen_name": "TwitterDev",
    "location": "Internet",
    "url": "https:\/\/dev.twitter.com\/",
    "description": "Your official source for Twitter Platform news, updates & events. 
    Need technical help? Visit https:\/\/twittercommunity.com\/ \u2328\ufe0f 
    #TapIntoTwitter"
  },
  "place": {   
  },
  "entities": {
    "hashtags": [      
    ],
    "urls": [
      {
        "url": "https:\/\/t.co\/XweGngmxlP",
        "unwound": {
          "url": "https:\/\/cards.twitter.com\/cards\/18ce53wgo4h\/3xo1c",
          "title": "Building the Future of the Twitter API Platform"
        }
      }
    ],
    "user_mentions": [     
    ]
  }
}
```

@@

LDA model fitting requires that the corpus be organized as a 
document-term matrix. In a document-term matrix, each row corresponds to a single document (a single 
tweet), and each column is a single term (or word). Each cell contains a 
count (the number of 
occurrences of a term in the specified document). We created a document-term matrix with the R function `cast_dtm` from the `tidytext` package.

@@

We identified the top ten most probable terms for each of ten topics in our models 
(Figures \@ref(fig:may24), \@ref(fig:may25), \@ref(fig:may26)). 
We plotted the within-topic word probabilities as bar graphs. We see that topic-specific 
word probabilities seldom exceed 0.05. We also note that some words are heavily weighted in 
multiple topics. This observation complicates semantic topic interpretation. We 
also caution that the results display expletives (that appeared on Twitter) and may be "not suitable for work (NSFW)".

@@

```{r, may25, fig.cap = "Top terms for LDA model from May 25, 2020 (Memorial Day). Results contain expletives and may be 'not suitable for work (NSFW)'."}
knitr::include_graphics("../results/beta-2020-05-25.png")
```

### 2020-11-02 Rmd/jse-special-issue-2020-october-anon.Rmd
added fixes for November 2020 typos

---
title: "What is happening on Twitter? A framework for student research projects with tweets"
author: "Frederick J. Boehm and Bret M. Hanlon"
date: ""
params:
  lastmod: !r lubridate::now()
bibliography: jse.bib
csl: american-statistical-association.csl
geometry: "left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm"
output: 
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: FALSE
    number_sections: TRUE
    keep_tex: TRUE
    includes:
      in_header: head.tex
  bookdown::word_document2:
    toc: FALSE
---

>>>>

---
title: "What is happening on Twitter? A framework for student research projects with tweets"
author: ""
date: ""
params:
  lastmod: !r lubridate::now()
bibliography: jse.bib
csl: american-statistical-association.csl
geometry: "left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm"
output: 
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: FALSE
    number_sections: TRUE
    keep_tex: TRUE
    includes:
      in_header: head.tex
  bookdown::word_document2:
    toc: FALSE
---
